\large How resource abundance and stochasticity affect animals' space-use requirements

### 2024-03-28 writing/appendix-C-empirical-modeling.Rmd
updated appendices and methods for empirical model

\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e. its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary. `mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2023] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs. Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families.

@@

\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the linear transformation

\begin{equation}
\nu_{[0, 1]} = \frac{\nu + 1}{2},
(\#eq:ndvi-to-01)
\end{equation}

\noindent where $\nu$ indicates the NDVI value in $[-1, 1]$ and $\nu_{[0, 1]}$ is the scaled NDVI value [see @denny_fallacy_2017 for more information on transforming response data]. <!--Note that only linear transformations (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$. We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is

@@

\noindent We can easily convert $\mu(t)$ and $\sigma^2(t)$ back to mean and variance in NDVI using the inverse of equation \@ref(eq:ndvi-to-01):

\singlespace <!-- to avoid excessive spacing -->
\begin{equation}
\nu = 2 \nu_{[0, 1]} - 1.
(\#eq:01-to-ndvi)
\end{equation}

\noindent Consequently, we have

\begin{equation}
\e(\nu|t) = 2 \mu(t) - 1.
(\#eq:mu-to-e-ndvi)
\end{equation}

\noindent and

\begin{equation}
\var(\nu|t) = 2^2 \sigma^2(t) + 0 = 4 \sigma^2(t).
(\#eq:sigma2-to-v-ndvi)
\end{equation}

@@

Particular attention should be be given when deciding what distribution to use and how to estimate means and variances in $R$. Improper models and simulations of resource abundance can fail to produce robust, sensible, and accurate estimates of $R$.

@@

This section of the appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript (fig. 5). The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs). Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig)).

@@

```{r, eval=FALSE}
getwd() # change working directory or file paths if needed
```

```{r, eval=TRUE, echo=FALSE}
print('H:/GitHub/hr-resource-stoch/writing') # for anonymity in peer review
```

```{r, eval=TRUE}
# attach all necessary packages
library('terra')     # to import and save rasters
library('dplyr')     # for data wrangling
library('purrr')     # for functional programming
library('tidyr')     # for data wrangling
library('ggplot2')   # for fancy plots
library('cowplot')   # for fancy multi-panel plots
library('ctmm')      # for movement modeling
library('mgcv')      # for empirical Bayesian GAMs
library('scam')      # for shape-constrained splines
library('lubridate') # for smoother date wrangling
library('sf')        # for spatial features
library('MODIStsp')  # for downloading NDVI rasters
library('dagitty')   # for directed acyclical graphs
library('ggdag')     # for directed acyclical graphs
theme_set(theme_bw()) # change default theme

# attach the beta location-scale family written by Simon Wood
source('../functions/betals.r')

# source all necessary scripts
source('../analysis/figures/default-figure-styling.R') # for color palettes
source('../earthdata-login-info.R') # personal login info for EarthData
source('../functions/window_hr.R') # function to calculate HRs
```

@@

# download NDVI rasters (if needed, create all necessary folders first)
MODIStsp(gui = FALSE, # do not use the browser GUI, only run in R
         out_folder = 'data/ndvi-rasters/tapir-anna',
         selprod = 'Vegetation Indexes_16Days_250m (M*D13Q1)',
         prod_version = '061', # 2022 raster version
         bandsel = 'NDVI', # NDVI layer only
         sensor = 'Terra', # only terrestrial values, ignore water
         user = USERNAME, # Earthdata username for urs.earthdata.nasa.gov
         password = PASSWORD, # your Earthdata password
         start_date = format(min(anna_tel$timestamp) - 16, '%Y.%m.%d'),
         end_date = format(max(anna_tel$timestamp) + 16, '%Y.%m.%d'),
         spatmeth = 'bbox', # use a bounding box for the extent
         bbox = bbox, # spatial file for raster extent
         out_projsel = 'User Defined', # use specified projection
         output_proj = '+proj=longlat', # download unprojected raster
         resampling = 'bilinear', # raster resampling method for new proj
         delete_hdf = TRUE, # delete HDF files after download is complete
         scale_val = TRUE, # convert from integers to floats within [-1, 1]
         out_format = 'GTiff', # output format
         verbose = TRUE) # print processing messages

# save NDVI data as an rds file of a tibble
list.files(path = 'data/ndvi-rasters/tapir-anna/VI_16Days_250m_v61/NDVI/',
           pattern = '.tif', full.names = TRUE) %>%
  rast() %>% # import all rasters as a single stack
  as.data.frame(xy = TRUE) %>% # convert to a data frame
  pivot_longer(-c(x, y)) %>% # change to long format (x, y, name, value)
  transmute(long = x, # rename x column
            lat = y, # rename y column
            date = substr(name, # change name to a date
                          start = nchar('MOD13Q1_NDVI_x'),
                          stop = nchar(name)) %>%
              as.Date(format = '%Y_%j'), # format is year_julian date
            ndvi = value, # rename value column
            dec_date = decimal_date(date)) %>%
  saveRDS('data/ndvi-rasters/tapir-anna/tapir-anna-data.rds')

@@

We removed the raster for 2017-12-19 because the values were unusually low for the region. We hypothesize the change in NDVI was drastic, temporary, and widespread because of a sudden flood (which is common for the Cerrado):

@@

```{r}
m_ndvi <-
  gam(list(
    # mean predictor
    ndvi ~ # not scaling because range is in (0, 1)
      s(long, lat, bs = 'ds', k = 20) + # mean over space
      s(dec_date, bs = 'tp', k = 20),   # high k to account for adaptions
    # scale predictor (sigma2 = mu * (1 - mu) * scale)
    ~
      s(long, lat, bs = 'ds', k = 10) + # scale over space
      s(dec_date, bs = 'tp', k = 10)),  # scale over time
    family = betals(),
    data = anna_ndvi,
    method = 'REML') # REstricted Maximum Likelihood 
```

Note that when fitting location-scale GAMs, one should pay particular attention to the number of knots used for each smooth term. While using a penalized maximum likelihood method such as REML (`method = 'REML'`) helps avoid over-fitting the model, finding the right balance between each of the `k` values is crucial. Excessively smooth terms for the mean can inflate the scale term (fig. C\@ref(fig:undersmooth)), while excessively wiggly terms for the mean can cause the scale to be under-estimated (fig. C\@ref(fig:oversmooth)). Ultimately, each of the `k` values should be decided in such a way as to mimic the animal's responsiveness, adaptability, motility, and memory (or ability to predict cycles or events). If one is unsure where to start from, keeping the `k` for the scale terms below half the `k` for the mean terms is a good starting point.

@@

```{r gamls-terms, echo=FALSE, eval=TRUE, fig.height=8, fig.cap="Estimated spatiotemporal trends in mean and scale parameters using the model detailed in the code chunk above. Estimates are provided on the logit link scale. The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms. Dashed lines inticate the 95\\% credible intervals for the temporal terms.", fig.pos='!h'}
m_ndvi <- readRDS('../models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
plot(m_ndvi, pages = 1, scheme = 3, n = 250, scale = 0)
```

```{r}
gam(list(
  # mean predictor (default ks are too small)
  ndvi ~ s(long, lat, bs = 'ds', k = 10) + s(dec_date, bs = 'tp', k = 10),
  # scale predictor (excessively high k)
  ~ s(long, lat, bs = 'ds', k = 20) + s(dec_date, bs = 'tp', k = 20)),
  family = betals(), data = anna_ndvi, method = 'REML') %>%
  plot(pages = 1, scheme = 3, n = 250, scale = 0)
```

```{r undersmooth, fig.cap="Estimated mean and scale parameters with excessively small basis size $\\tt{k}$ for the mean's smooth terms and excessively large for the scale's smooth terms. The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms. Dashed lines inticate the 95\\% credible intervals for the temporal terms.", eval=TRUE, fig.height=6, fig.pos='!h', echo=FALSE}
gam(list(
  # mean predictor (default ks are too small)
  ndvi ~ s(long, lat, bs = 'ds', k = 10) + s(dec_date, bs = 'tp', k = 10),
  # scale predictor (excessively high k)
  ~ s(long, lat, bs = 'ds', k = 20) + s(dec_date, bs = 'tp', k = 20)),
  family = betals(), data = anna_ndvi, method = 'REML') %>%
  plot(pages = 1, scheme = 3, n = 250, scale = 0)
```

```{r}
gam(list(
  # mean predictor (ks too large)
  ndvi ~ s(long, lat, bs = 'ds', k = 50) +
    s(dec_date, bs = 'tp', k = n_distinct(anna_ndvi$dec_date)),
  # scale predictor (ks too small)
  ~ s(long, lat, bs = 'ds', k = 5) + s(dec_date, bs = 'tp', k = 5)),
  family = betals(), data = anna_ndvi, method = 'REML') %>%
  plot(pages = 1, scheme = 3, n = 250, scale = 0)
```

```{r oversmooth, fig.cap="Estimated mean and scale parameters in NDVI with excessively large basis size $\\tt{k}$ for the smooth term of time ($\\tt{dec\\_date}$) for the mean in NDVI and excessively low for the variance temporal smooth term. The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms. Dashed lines inticate the 95\\% credible intervals for the temporal terms.", eval=TRUE, fig.height=6, fig.pos='!h', echo=FALSE}
gam(list(
  # mean predictor (ks too large)
  ndvi ~ s(long, lat, bs = 'ds', k = 50) +
    s(dec_date, bs = 'tp', k = n_distinct(anna_ndvi$dec_date)),
  # scale predictor (ks too small)
  ~ s(long, lat, bs = 'ds', k = 5) + s(dec_date, bs = 'tp', k = 5)),
  family = betals(), data = anna_ndvi, method = 'REML') %>%
  plot(pages = 1, scheme = 3, n = 250, scale = 0)
```

@@

Next, we can estimate the mean and variance in NDVI for each 7-day period by taking the average for the GPS locations within each period. We also add a column of weights that are based on the home range confidence estimates' 95% intervals.

```{r}
tapir <-
  readRDS('../models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
  mutate(est = map(dataset,
                   \(.d) filter(anna_tel, timestamp %in% .d$timestamp)),
         mu = map_dbl(est, \(.d) mean(.d$mu)),
         sigma2 = map_dbl(est, \(.d) mean(.d$sigma2))) %>%
  select(date, mu, sigma2, hr_lwr_95, hr_est_95, hr_upr_95) %>%
  mutate(weight = 1 / (hr_upr_95 - hr_lwr_95), # small CI => large weight
         weight = weight / mean(weight)) # standardize weights
sum(tapir$weight) == nrow(tapir) # ensure weights are right
```

```{r, eval=TRUE, echo=FALSE}
tapir <- readRDS('../data/anna-hr-ndvi-data.rds')
sum(tapir$weight) == nrow(tapir) # ensure weights are right
```

@@

```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
theme_set(theme_bw()) # need to set the theme again for some reason
# axis labels (spaces in hr_lab to avoid overlap with the plot label)
# unicode characters removed because they knitting errors
e_r <- bquote(paste(bold('Resource abundance, E('), bolditalic(R),
                    bold(')')))
v_r <- bquote(paste(bold('Resource stochasticity, Var('), bolditalic(R),
                    bold(')')))
hr_lab <- bquote(bold(paste('7-day home range size (', km^2, ')   ')))

l_grobs <- lapply(
  list(ggplot(tapir, aes(date, mu)) + # mean
         geom_line(color = pal[1], size = 2) +
         labs(x = NULL, y = e_r),
       ggplot(tapir, aes(date, sigma2)) + # variance
         geom_line(color = pal[2], size = 2) +
         labs(x = NULL, y = v_r),
       ggplot(tapir, aes(date, hr_est_95)) + # 95% home range
         geom_line(color = pal[3], size = 2) +
         labs(x = NULL, y = hr_lab)),
  as_grob) # convert to grid graphical objects (grobs)

@@

<!-- \clearpage -->

@@

```{r dag, eval=TRUE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$, where NDVI was used as a proxy for $R$. $Z$ and $S$ indicate unaccounted confounding factors that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).", fig.height=4, fig.width=4}
dag <- dagitty('dag{
Z -> R -> "NDVI" <- S <- R
R -> H <- Z
}')
coordinates(dag) <- list(x = c(R = 0, 'NDVI' = 0, Z = 1, S = 1, H = 0),
                         y = c(R = 0, 'NDVI' = 1, Z = 0, S = 1, H = -1))
plot(dag)
```

@@

```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space use (on the log link scale) using a GAM with no shape constraints. The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels. Shaded areas inticate the 95\\% credible intervals."}
gam(hr_est_95 ~ s(mu, bs = 'ts', k = 5) + s(sigma2, bs = 'ts', k = 5),
    family = Gamma('log'),
    weights = weight, # weights based on CI size
    data = tapir,
    method = 'REML') %>%
  plot(pages = 1, scheme = 1) # trends should be monotonic
```

Due to the small sample size, the spatiotemporal autocorrelation, and the correlation between $\mu(t)$ and $\sigma^2(t)$, it is hard to separate the effects of the two predictors. However, since it is reasonable to expect the effects of $\mu(t)$ and $\sigma^2(t)$ to be monotonic decreaseing and increasing, respectively, we can use a Shape-Constrained Additive Model (SCAM) via the `scam` package [@pya_scam_2022].

```{r hr-scam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space use (on the log link scale) using a shape-constrained GAM. The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels. Shaded areas inticate the 95\\% credible intervals."}
m <- scam(hr_est_95 ~
            s(mu, bs = 'mpd', k = 5) +
            s(sigma2, bs = 'mpi', k = 5),
          family = Gamma('log'),
          weights = weight,
          data = tapir)
plot(m, pages = 1, scheme = 1)
```

This SCAM fits substantially better than a SCAM that only accounts for $\mu(t)$:

```{r, eval=TRUE}
m_mu <- scam(hr_est_95 ~ s(mu, bs = 'mpd', k = 5),
             family = Gamma('log'),
             weights = weight,
             data = tapir)
round(AIC(m, m_mu), 3)
```

We can now create the predictions from the shape-constrained GAM.

```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
theme_set(theme_bw()) # need to set the theme again for some reason
# predict marginal effects for each term
preds <- tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
                sigma2 = gratia:::seq_min_max(tapir$sigma2, n = 250)) %>%
  bind_cols(
    # predictions for the marginal effect of mu
    predict(m, newdata = ., terms = 's(mu)', type = 'link',
            se.fit = TRUE) %>%
      as.data.frame() %>% # convert list to data frame
      transmute(hr_mu_est = exp(fit), # transform to the response scale
                hr_mu_lwr = exp(fit - 1.96 * se.fit), # 95% cred. intervals
                hr_mu_upr = exp(fit + 1.96 * se.fit)),
    # predictions for the marginal effect of sigma2
    predict(m, newdata = ., terms = 's(sigma2)', type = 'link',
            se.fit = TRUE) %>%
      as.data.frame() %>%
      transmute(hr_sigma2_est = exp(fit),
                hr_sigma2_lwr = exp(fit - 1.96 * se.fit),
                hr_sigma2_upr = exp(fit + 1.96 * se.fit)))

p_d <- ggplot() +
  coord_cartesian(ylim = c(0, 12.5)) +
  geom_point(aes(mu, hr_est_95), tapir, color = pal[3], alpha = 0.5) +
  geom_ribbon(aes(mu, ymin = hr_mu_lwr, ymax = hr_mu_upr), preds,
              fill = pal[1], alpha = 0.3) +
  geom_line(aes(mu, hr_mu_est), preds, color = pal[1], linewidth = 2) +
  xlab(e_r) +
  scale_y_continuous(hr_lab)

p_e <- ggplot() +
  coord_cartesian(ylim = c(0, 12.5)) +
  geom_point(aes(sigma2, hr_est_95), tapir, color = pal[3], alpha = 0.5) +
  geom_ribbon(aes(sigma2, ymin = hr_sigma2_lwr, ymax = hr_sigma2_upr),
              preds, fill = pal[2], alpha = 0.3) +
  geom_line(aes(sigma2, hr_sigma2_est), preds, color = pal[2],
            linewidth = 2) +
  xlab(v_r) +
  scale_y_continuous(hr_lab)

p_f <- ggplot(tapir) +
  geom_point(aes(mu, sigma2), alpha = 0.5) +
  labs(x = e_r, y = v_r)

r_grobs <- map(list(p_d, p_e, p_f), as_grob)

# align right margins of all plots
aligned_widths <- align_margin(map(r_grobs, \(x) {x$widths}), 'first')

# Setting the dimensions of plots to the aligned dimensions
for(i in seq_along(r_grobs)) r_grobs[[i]]$widths <- aligned_widths[[i]]

>>>>

\noindent Location-scale models [theory: @rigby_generalized_2005; @stasinopoulos_generalized_2007; examples: @bjorndahl_abrupt_2022; @marien_timing_2022; @gushulak_impacts_2024] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e. its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary. `mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2023] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs. Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families.

@@

\noindent Although there is no commonly-used distribution with a support over the full range of NDVI (i.e., $[-1, 1]$), we use beta distribution (support (0, 1)) since the all NDVI values were sufficiently greater than 0 (range: 0.3534 to 0.9475). Thus, here we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$. We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is

\singlespace

@@

@@

Particular attention should be be given when deciding what distribution to use and how to estimate means and variances in $R$. Improper models and simulations of resource abundance can fail to produce robust, sensible, and accurate estimates of $R$. The following section addresses these concerns.

@@

This section of the appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript (fig. 5). The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs). Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and space use (section \@ref(final-fig)).

@@

```{r, eval=TRUE, warning=FALSE}
# attach all necessary packages
library('terra')     # to import and save rasters
library('dplyr')     # for data wrangling
library('purrr')     # for functional programming
library('tidyr')     # for data wrangling
library('ggplot2')   # for fancy plots
library('cowplot')   # for fancy multi-panel plots
library('ctmm')      # for movement modeling
library('mgcv')      # for empirical Bayesian GAMs
library('lubridate') # for smoother date wrangling
library('sf')        # for spatial features
library('MODIStsp')  # for downloading NDVI rasters
library('dagitty')   # for directed acyclical graphs
library('ggdag')     # for directed acyclical graphs
library('gratia')    # for ggplot-based GAM figures
theme_set(theme_bw()) # change default theme
source('../functions/betals.r') # betals family written by Simon Wood
source('../analysis/figures/default-figure-styling.R') # for color palettes
source('../earthdata-login-info.R') # personal login info for EarthData
source('../functions/window_hr.R') # function to calculate HRs
```

@@

# download NDVI rasters (if needed, create all necessary folders first)
MODIStsp(gui = FALSE, # do not use the browser GUI, only run in R
         out_folder = 'data/ndvi-rasters/tapir-anna',
         selprod = 'Vegetation Indexes_16Days_250m (M*D13Q1)',
         prod_version = '061', # 2022 raster version
         bandsel = 'NDVI', # NDVI layer only
         sensor = 'Terra', # only terrestrial values, ignore water
         user = USERNAME, # Earthdata username for urs.earthdata.nasa.gov
         password = PASSWORD, # your Earthdata password
         start_date = format(min(anna_tel$timestamp) - 16, '%Y.%m.%d'),
         end_date = format(max(anna_tel$timestamp) + 16, '%Y.%m.%d'),
         spatmeth = 'bbox', # use a bounding box for the extent
         bbox = bbox, # spatial file for raster extent
         out_projsel = 'User Defined', # use specified projection
         output_proj = '+proj=longlat', # download unprojected raster
         resampling = 'bilinear', # raster resampling method for new proj
         delete_hdf = TRUE, # delete HDF files after download is complete
         scale_val = TRUE, # convert from integers to floats within [-1, 1]
         out_format = 'GTiff', # output format
         verbose = TRUE) # print processing messages
```

```{r}
# save NDVI data as an rds file of a tibble
list.files(path = 'data/ndvi-rasters/tapir-anna/VI_16Days_250m_v61/NDVI/',
           pattern = '.tif', full.names = TRUE) %>%
  rast() %>% # import all rasters as a single stack
  as.data.frame(xy = TRUE) %>% # convert to a data frame
  pivot_longer(-c(x, y)) %>% # change to long format (x, y, name, value)
  transmute(long = x, # rename x column
            lat = y, # rename y column
            date = substr(name, # change name to a date
                          start = nchar('MOD13Q1_NDVI_x'),
                          stop = nchar(name)) %>%
              as.Date(format = '%Y_%j'), # format is year_julian date
            ndvi = value, # rename value column
            dec_date = decimal_date(date)) %>%
  saveRDS('data/ndvi-rasters/tapir-anna/tapir-anna-data.rds')

@@

We removed the raster for 2017-12-19 because the values were unusually low for the region. We hypothesize the change in NDVI was drastic, temporary, and widespread because of a sudden flood. While sudden floods are common for the Cerrado, we believe NDVI was not representative of the available forage availability.

@@

```{r}
m_ndvi <-
  gam(list(
    # mean predictor
    ndvi ~ # not scaling because range is in (0, 1)
      s(long, lat, bs = 'ds', k = 50) + # mean over space
      s(dec_date, bs = 'tp', k = 10), # mean over time
    # scale predictor (sigma2 = mu * (1 - mu) * scale)
    ~
      s(long, lat, bs = 'ds', k = 30) + # scale over space
      s(dec_date, bs = 'tp', k = 10)),  # scale over time
    family = betals(),
    data = anna_ndvi,
    method = 'REML') # REstricted Maximum Likelihood 
```

Note that when fitting location-scale GAMs, one should pay particular attention to the number of knots used for each smooth term. While using a penalized maximum likelihood method such as REML (`method = 'REML'`) helps avoid over-fitting the model, it does not prevent over-fitting, so finding the right balance between each of the `k` values is crucial. Excessively smooth terms for the mean can inflate the scale term, while excessively wiggly terms for the mean can cause the scale to be under-estimated. Ultimately, each of the `k` values should be decided in such a way as to mimic the animal's responsiveness, adaptability, motility, and memory (or ability to predict cycles or events). If one is unsure where to start from, keeping the `k` for the scale terms below half the `k` for the mean terms is a good starting point.

@@

```{r gamls-terms, echo=FALSE, eval=TRUE, fig.height=8, fig.cap="Estimated spatiotemporal trends in mean and scale parameters using the model detailed in the code chunk above. Estimates are provided on the logit link scale. The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms. Shaded ribbons inticate the 95\\% credible intervals for the temporal terms.", fig.height=6, fig.pos='!h'}
m_ndvi <- readRDS('../models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
draw(m_ndvi, rug = FALSE) &
  theme_bw() + theme(panel.grid = element_blank())
```

@@

Next, we can estimate the mean and variance in NDVI for each 7-day period by taking the average for the GPS locations within each period.

```{r}
tapir <-
  readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
  mutate(sub_tel = map(dataset,
                       \(.d) filter(tel, timestamp %in% .d$timestamp)),
         mu = map_dbl(sub_tel, \(.d) mean(.d$mu)),
         sigma2 = map_dbl(sub_tel, \(.d) mean(.d$sigma2))) %>%
  select(date, mu, sigma2, hr_est_95)
```

@@

```{r, eval=TRUE, echo=FALSE}
tapir <- readRDS('../data/anna-hr-ndvi-data.rds')
```

```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
# need to set the theme again for some reason
theme_set(theme_bw() + theme(panel.grid = element_blank()))

# axis labels (unicode characters removed because they knitting errors)
e_r <- bquote(paste(bold('Resource abundance')))
v_r <- bquote(paste(bold('Resource stochasticity')))
hr_lab <- bquote(bold(paste('7-day space-use requirement (', km^2, ')')))

l_grobs <- lapply(
  list(ggplot(tapir, aes(date, mu)) + # mean
         geom_line(color = pal[1], linewidth = 2) +
         labs(x = NULL, y = e_r),
       ggplot(tapir, aes(date, sigma2)) + # variance
         geom_line(color = pal[2], linewidth = 2) +
         labs(x = NULL, y = v_r),
       ggplot(tapir, aes(date, hr_est_95)) + # 95% home range
         geom_line(color = pal[3], linewidth = 2) +
         labs(x = NULL, y = hr_lab)),
  as_grob) # convert to grid graphical objects (grobs)

@@

@@

```{r dag, eval=TRUE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$, where NDVI was used as a proxy for $R$. $Z$ and $S$ indicate unaccounted confounding factors that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).", fig.height=4, fig.width=6, out.width='50%'}
dag <- dagitty('dag{
Z -> R -> "NDVI" <- S <- R
R -> H <- Z
}')
coordinates(dag) <- list(x = c(R = 0, 'NDVI' = 0, Z = 1, S = 1, H = 0),
                         y = c(R = 0, 'NDVI' = 1, Z = 0, S = 1, H = -1))
ggdag(dag) + theme_dag()
```

@@

```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space use (on the log link scale). The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels. Shaded areas inticate the 95\\% credible intervals.", fig.height=3}
m <- gam(hr_est_95 ~ s(mu, k = 4) + s(sigma2, k = 4),
         family = Gamma('log'), data = tapir, method = 'REML')
draw(m) & theme_bw() + theme(panel.grid = element_blank())
```

We can now create the predictions from the GAM.

```{r, eval=TRUE, out.width='\\linewidth'}
# predict marginal effects for each term
preds <- tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
                sigma2 = seq(5e-4, 25e-4, length.out = 250)) %>%
  bind_cols(
    # predictions for the marginal effect of mu
    predict(m, newdata = ., exclude = 's(sigma2)',
            type = 'link', se.fit = TRUE, unconditional = TRUE) %>%
      as.data.frame() %>%
      transmute(hr_mu_est = exp(fit),
                hr_mu_lwr = exp(fit - 1.96 * se.fit),
                hr_mu_upr = exp(fit + 1.96 * se.fit)),
    # predictions for the marginal effect of sigma2
    predict(m, newdata = ., exclude = 's(mu)',
            type = 'link', se.fit = TRUE, unconditional = TRUE) %>%
      as.data.frame() %>%
      transmute(hr_sigma2_est = exp(fit),
                hr_sigma2_lwr = exp(fit - 1.96 * se.fit),
                hr_sigma2_upr = exp(fit + 1.96 * se.fit)))
```

```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
theme_set(theme_bw() + theme(panel.grid = element_blank()))

p_d <- ggplot() +
  coord_cartesian(ylim = c(0, 12.5)) +
  geom_point(aes(mu, hr_est_95), tapir, alpha = 0.3, color = pal[3]) +
  geom_ribbon(aes(mu, ymin = hr_mu_lwr, ymax = hr_mu_upr), preds,
              fill = pal[1], alpha = 0.3) +
  geom_line(aes(mu, hr_mu_est), preds, color = pal[1], linewidth = 2) +
  labs(x = e_r, y = hr_lab)

p_e <- ggplot() +
  geom_point(aes(sigma2, hr_est_95), tapir, alpha = 0.3, color = pal[3]) +
  geom_ribbon(aes(sigma2, ymin = hr_sigma2_lwr, ymax = hr_sigma2_upr),
              preds, fill = pal[2], alpha = 0.3) +
  geom_line(aes(sigma2, hr_sigma2_est), preds, color = pal[2],
            linewidth = 2) +
  labs(x = v_r, y = hr_lab)

p_f <-
  expand_grid(mu = seq(from = floor(min(tapir$mu) * 100) / 100,
                       to = ceiling(max(tapir$mu) * 100) / 100,
                       length.out = 250),
              sigma2 = seq(from = 0, to = 0.0035, length.out = 250)) %>%
  mutate(hr_full_est = predict(m, newdata = ., type = 'response')) %>%
  ggplot() +
  geom_raster(aes(mu, sigma2, fill = hr_full_est)) +
  geom_contour(aes(mu, sigma2, z = hr_full_est), color = 'black') +
  geom_point(aes(mu, sigma2), tapir, alpha = 0.3, show.legend = FALSE) +
  scale_x_continuous(e_r, expand = c(0, 0)) +
  scale_y_continuous(v_r, expand = c(0, 0)) +
  scale_fill_gradient(bquote(atop(bold('7-day space-use'),
                                  paste(bold('requirement (km'), '\U00B2',
                                        bold(')')))),
                      low = 'grey90', high = pal[3], limits = c(0, NA)) +
  theme(legend.position = c(1, 1),
        legend.justification = c('right', 'top'),
        legend.box.background = element_rect(),
        legend.background = element_rect(),
        legend.key.width = unit(0.35, 'in')) +
  guides(fill = guide_colorbar(
    title.position = 'top',
    theme = theme(legend.title = element_text(hjust = 1)),
    direction = 'horizontal'))

# align right margins of all plots
r_grobs <- map(list(p_d, p_e, p_f), as_grob)
aligned_widths <- align_margin(map(r_grobs, \(x) {x$widths}), 'first')
for(i in seq_along(r_grobs)) r_grobs[[i]]$widths <- aligned_widths[[i]]
